---
title: "P8105_hw3_yc3577"
author: "Yimeng Cai"
date: "10/12/2023"
output: github_document
---

```{r}
library(p8105.datasets)
library(tidyverse)
data("instacart")

instacart = instacart |> as_tibble()
```
This dataset contains 1384617 rows and 15 columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are 39123 products found in 131209 orders from 131209 distinct users.
Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart |>
  count(aisle) |>
  arrange(n, desc = FALSE)
```
```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```

Problem 2

```{r}
library(p8105.datasets)
data("brfss_smart2010")
```


1. format the data to use appropriate variable names;
2. focus on the “Overall Health” topic
3. include only responses from “Excellent” to “Poor”
4. organize responses as a factor taking levels ordered from “Poor” to “Excellent”

```{r}
brfss = 
  brfss_smart2010 |>
  as_tibble()

brfss = 
  brfss |>
  janitor :: clean_names() |>
  rename(state = locationabbr, county = locationdesc) |>
  filter(topic %in% 'Overall Health') |>
  mutate(response = fct_relevel(response, 
                                'Poor', 'Fair', 'Good', 'Very good', 'Excellent')) |>
  filter(response %in% c('Poor', 'Excellent'))

brfss
```


In 2002, which states were observed at 7 or more locations?
```{r}
brfss_2002 =
  brfss |>
  filter(year == '2002') |>
  group_by(year, state) |>
  count(state) |>
  filter( n >= 7) |>
  pivot_wider(
    names_from = year, 
    values_from = state)

brfss_2002
```

What about in 2010?
```{r}
brfss_2010 =
  brfss |>
  filter(year == '2010') |>
  group_by(year, state) |>
  count(state) |>
  filter( n >= 7) |>
  pivot_wider(
    names_from = year, 
    values_from = state)

brfss_2010
```

Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. 
```{r}
brfss_dataset =
  brfss |>
  filter(response == 'Excellent') |>
  select(year, state, data_value) |>
  group_by(year, state) |>
  summarize(mean_dv = mean(data_value))

brfss_dataset
```

Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).
```{r}
brfss_spaghetti =
  brfss_dataset |>
  ggplot(aes(x = year, y = mean_dv, color = state, group = state)) + 
  geom_line() +
  labs(
    title = 'Mean Data Value over Time',
    x = 'Year',
    y = 'Average Data Value',
  )

brfss_spaghetti
```

Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.
```{r}

brfss_box = 
  brfss |>
  filter(state %in% 'NY') |>
  filter(year %in% c('2006','2010')) |>
  ggplot(aes(x = response, y = data_value)) +
  geom_boxplot() +
  facet_grid(. ~ year) +
  labs(
    title = 'Data Value distribution over Response in NY for 2006 & 2010',
    x = 'Response',
    y = 'Data Value'
  )

brfss_box
```



Problem 3


Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).
```{r}
nhanes_covar =
  read_csv('nhanes_covar.csv', skip = 4) |>
  janitor::clean_names() |>
  filter(age > 21) |>
  mutate(
    sex = recode(sex, 
                 '1' = 'male',
                 '2' = 'female'),
    education = recode(education,
                       '1' = 'Less than high school',
                       '2' = 'High school equivalent',
                       '3' = 'More than high school'),
    sex = as.factor(sex),
    education = as.factor(education)) |>
  drop_na()

nhanes_covar
```
```{r}
nhanes_accel = 
  read_csv('nhanes_accel.csv') |>
  janitor::clean_names() |>
  pivot_longer(
    min1:min1440,
    values_to = 'MIMS',
    names_to = 'Number',
    names_prefix = 'min') |>
  drop_na()

nhanes_accel
  
```


Joined the two table together. 
```{r}
nhanes_joint = 
  left_join(nhanes_covar, nhanes_accel, by = 'seqn')

nhanes_joint
```


Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.
```{r}
nhanes_table =
  nhanes_joint |>
  group_by(sex, education) 


age_plot = 
  nhanes_table |>
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(alpha = 0.5) +
  facet_grid(.~education)

age_plot
```


Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r}
nhanes_clean =
  nhanes_table |>
  group_by(seqn) |>
  mutate(
    total_activity = sum(MIMS)
  )

nhanes_clean

activity_plot =
  nhanes_clean |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(.~education) +
  labs(
    title = 'Total activities over Age',
    x = 'Age',
    y = 'Total Activities'
  )

activity_plot
```


Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.
```{r}
threepanel_plot =
    nhanes_table |>
  ggplot(aes(x = Number, y = MIMS, group = seqn, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(aes(group = sex), se = FALSE) +
  facet_grid(.~education) +
  labs(
    title = '24-Hour Activities over Age for Education Level',
    x = 'Time',
    y = 'Total Activities'
  )

threepanel_plot

  
```

