---
title: "P8105_hw3_yc3577"
author: "Yimeng Cai"
date: "10/12/2023"
output: github_document
---

```{r}
library(p8105.datasets)
library(tidyverse)
data("instacart")

instacart = instacart |> as_tibble()
```
This dataset contains 1384617 rows and 15 columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are 39123 products found in 131209 orders from 131209 distinct users.
Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart |>
  count(aisle) |>
  arrange(desc(n))
```
Next is a plot that shows the number of items ordered in each aisle.
Here, aisles are ordered by ascending number of items.
```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
Our next table shows the three most popular items in aisles
`baking ingredients`, `dog food care`, and `packaged vegetables fruits`,
and includes the number of times each item is ordered in your table.
```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady
Apples and Coffee Ice Cream are ordered on each day of the week. This
table has been formatted in an untidy manner for human readers. Pink
Lady Apples are generally purchased slightly earlier in the day than
Coffee Ice Cream, with the exception of day 5.
```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)

    ## `summarise()` has grouped output by 'product_name'. You can override using the
    ## `.groups` argument.
```

Problem 2

First, we need to use `library` and `data` to get our `brfss_smart2010` data.
```{r}
library(p8105.datasets)
data("brfss_smart2010")
brfss_smart2010
```


1. format the data to use appropriate variable names;
2. focus on the “Overall Health” topic
3. include only responses from “Excellent” to “Poor”
4. organize responses as a factor taking levels ordered from “Poor” to “Excellent”

We build the tibble with `as_tibble()` first to format the data in the way we want, and then we need to use `janitor::clean_names()` to format the variable names. 
Then we use the rename to simplify the variables like state and county by `rename(state = locationabbr, county = locationdesc)`. We use `filter` function to focus the topic on `Overall Health`. Finally we use the mutate function to build a new variable called response that only taking levels from poor to excellent with `mutate(response = fct_relevel(response, 'Poor', 'Fair', 'Good', 'Very good', 'Excellent'))`. 
```{r}
brfss = 
  brfss_smart2010 |>
  as_tibble()

brfss = 
  brfss |>
  janitor :: clean_names() |>
  rename(state = locationabbr, county = locationdesc) |>
  filter(topic %in% 'Overall Health') |>
  mutate(response = fct_relevel(response, 
                                'Poor', 'Fair', 'Good', 'Very good', 'Excellent'))

brfss
```


In 2002, which states were observed at 7 or more locations?
In 2002, CT, FL, MA, NC, NJ, PA these 6 states were observed with 7 or more locations.

To know which states had observed 7 or more locations in 2002, we firstly need to use filter to filter our the 2002 year rows with `filter(year == '2002')`, then, we use `group_by(state)` function to get each state grouped together. Then we need to count the numbers of individual distinct county in each states and put the number into a new columns of `loc_num` by `summarise()`. Finally, we `filter()` the states with `loc_number >=7` to get the results we want, and `knitr::` the table. 
```{r}
brfss_2002 =
  brfss |>
  filter(year == '2002') |>
  group_by(state) |>
  summarise(loc_num = n_distinct(county)) |>
  filter( loc_num >= 7) |>
  knitr::kable()

brfss_2002
```

What about in 2010?
In 2010, CA, CO, FL, MA, MD, NC, NE, NJ, NY, OH, PA, SC, TX, and WA these 14 states were observed with 7 or more locations.

To know which states had observed 7 or more locations in 2010, we firstly need to use filter to filter our the 2010 year rows with `filter(year == '2010')`, then, we use `group_by(state)` function to get each state grouped together. Then we need to count the numbers of individual distinct county in each states and put the number into a new columns of `loc_num` by `summarise()`. Finally, we `filter()` the states with `loc_number >=7` to get the results we want, and `knitr::` the table. 
```{r}
brfss_2010 =
  brfss |>
  filter(year == '2010') |>
  group_by(state) |>
  summarise( loc_num = n_distinct(county))|>
  filter( loc_num >= 7) |>
  knitr::kable()

brfss_2010
```

Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. 

To construct the dataset that is limited to excellent responses, we use `filter()` function to focus on `response == 'Excellent'`. Then we use `select()` function to select the columns only with year, state, and data_value variables. To make the dataset more clean and easy to observed, we will use `group_by(year, state)` to help us to see the values grouped in same year and individual states. Finally, we use `summarize()` function to get the construct a new column of `mean_dv` to get the mean of `data_value`. Then we are ready to make the graph.
```{r}
brfss_dataset =
  brfss |>
  filter(response == 'Excellent') |>
  select(year, state, data_value) |>
  group_by(year, state) |>
  summarize(mean_dv = mean(data_value))

brfss_dataset
```

Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

It's time to make a graph. To make a spaghetti plot, we need to use `ggplot(aes(x = year, y = mean_dv, color = state, group = state))` to take in the value of `x = year`, `y = mean_dv`, and specify that we need to have different states in different colors with `color = state` and `group = state`. Since it is a spaghetti graph, we need to use `geom_line()` to plot and add title names and x, y variable names using `labs(title = 'Mean Data Value over Time', x = 'Year', y = 'Average Data Value')`. Our spaghetti plot is created and is save to pdf using `ggsave()`.
```{r}
brfss_spaghetti =
  brfss_dataset |>
  ggplot(aes(x = year, y = mean_dv, color = state, group = state)) + 
  geom_line() +
  labs(
    title = 'Mean Data Value over Time',
    x = 'Year',
    y = 'Average Data Value')

brfss_spaghetti

ggsave('brfss_spaghetti.pdf', height = 10, width = 12)
```
Comments:
From the graph, we can see that most of the states have mean data_value range from 16.5 to 27.5, and most of them have up and down values between 2002 to 2010. Some states like CA and CO had relatively high mean dv all the time around 28, and some states like WV have an extreme low mean dv at 2005 and 2007 (below 15). However, though with variation between the 8 years, the overall change of mean data_value is relatively small from 2002 to 2010.

Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

To make a two-panel plot among locations in NY state, we use `filter(state %in% 'NY')` to get NY rows selected, then we `filter()` for the years 2006 and 2010, we need to use `filter(year %in% c('2006','2010'))`. 
Now we are ready to plot the two-panel plot with `ggplot(aes(x = response, y = data_value))` to take in the value of `x = response`, `y = data_value`. Then we need to use `geom_boxplot()` to plot the boxplot which is clear to comparison and use `facet_grid(. ~ year)` to wrap the 1d to 2d. In this case, we are comparing 2006 and 2010 years so we wrap the variable year to make a side by box-plot. Then we add title names and x, y variable names using `labs(title = 'Data Value distribution over Response in NY for 2006 & 2010', x = 'Response', y = 'Data Value')`. Our two-panel box-plot is created and is save to pdf using `ggsave()`.
```{r}

brfss_box = 
  brfss |>
  filter(state %in% 'NY') |>
  filter(year %in% c('2006','2010')) |>
  ggplot(aes(x = response, y = data_value)) +
  geom_boxplot() +
  facet_grid(. ~ year) +
  labs(
    title = 'Data Value distribution over Response in NY for 2006 & 2010',
    x = 'Response',
    y = 'Data Value'
  )

brfss_box

ggsave('brfss_box.pdf', height = 10, width = 12)
```
Comments:
From the graph, we can see that in the two years 2006 and 2010, the response patterns are similar that good and very good responses have relatively high data value around 30-35, and excellent response is in the middle around 20-25, then the fair around 10-15 and lastly the poor having lowest data value below 5. Difference is in 2010, the very good response obtaining higher data value of around 35 dv along with high variation (longer box) and the good response having lower data value of around 32 dv with low variation (shorter box). Excellent is relatively the same with longer low data value as a tail in 2010, and fair response having lower variation (shorter box + shorter tail). 


Problem 3

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).
```{r}
nhanes_covar =
  read_csv('nhanes_covar.csv', skip = 4) |>
  janitor::clean_names() |>
  filter(age > 21) |>
  mutate(
    sex = recode(sex, 
                 '1' = 'male',
                 '2' = 'female'),
    education = recode(education,
                       '1' = 'Less than high school',
                       '2' = 'High school equivalent',
                       '3' = 'More than high school'),
    sex = as.factor(sex),
    education = as.factor(education)) |>
  drop_na()

nhanes_covar
```
```{r}
nhanes_accel = 
  read_csv('nhanes_accel.csv') |>
  janitor::clean_names() |>
  pivot_longer(
    min1:min1440,
    values_to = 'MIMS',
    names_to = 'Number',
    names_prefix = 'min') |>
  drop_na()

nhanes_accel
  
```


Joined the two table together. 
```{r}
nhanes_joint = 
  left_join(nhanes_covar, nhanes_accel, by = 'seqn')

nhanes_joint
```


Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.
```{r}
nhanes_table =
  nhanes_joint |>
  group_by(sex, education) 


age_plot = 
  nhanes_table |>
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(alpha = 0.5) +
  facet_grid(.~education)

age_plot
```


Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r}
nhanes_clean =
  nhanes_table |>
  group_by(seqn) |>
  mutate(
    total_activity = sum(MIMS)
  )

nhanes_clean

activity_plot =
  nhanes_clean |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(.~education) +
  labs(
    title = 'Total activities over Age',
    x = 'Age',
    y = 'Total Activities'
  )

activity_plot
```


Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.
```{r}
threepanel_plot =
    nhanes_table |>
  ggplot(aes(x = Number, y = MIMS, group = seqn, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(aes(group = sex), se = FALSE) +
  facet_grid(.~education) +
  labs(
    title = '24-Hour Activities over Age for Education Level',
    x = 'Time',
    y = 'Total Activities'
  )

threepanel_plot

  
```

