---
title: "P8105_hw3_yc3577"
author: "Yimeng Cai"
date: "10/12/2023"
output: github_document
---

```{r}
library(p8105.datasets)
library(tidyverse)
data("instacart")

instacart = instacart |> as_tibble()
```
This dataset contains 1384617 rows and 15 columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are 39123 products found in 131209 orders from 131209 distinct users.
Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart |>
  count(aisle) |>
  arrange(desc(n))
```
Next is a plot that shows the number of items ordered in each aisle.
Here, aisles are ordered by ascending number of items.
```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
Our next table shows the three most popular items in aisles
`baking ingredients`, `dog food care`, and `packaged vegetables fruits`,
and includes the number of times each item is ordered in your table.
```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady
Apples and Coffee Ice Cream are ordered on each day of the week. This
table has been formatted in an untidy manner for human readers. Pink
Lady Apples are generally purchased slightly earlier in the day than
Coffee Ice Cream, with the exception of day 5.
```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)

    ## `summarise()` has grouped output by 'product_name'. You can override using the
    ## `.groups` argument.
```

Problem 2

First, we need to use `library` and `data` to get our `brfss_smart2010` data.
```{r}
library(p8105.datasets)
data("brfss_smart2010")
brfss_smart2010
```


1. format the data to use appropriate variable names;
2. focus on the “Overall Health” topic
3. include only responses from “Excellent” to “Poor”
4. organize responses as a factor taking levels ordered from “Poor” to “Excellent”

We build the tibble with `as_tibble()` first to format the data in the way we want, and then we need to use `janitor::clean_names()` to format the variable names. 
Then we use the rename to simplify the variables like state and county by `rename(state = locationabbr, county = locationdesc)`. We use `filter` function to focus the topic on `Overall Health`. Finally we use the mutate function to build a new variable called response that only taking levels from poor to excellent with `mutate(response = fct_relevel(response, 'Poor', 'Fair', 'Good', 'Very good', 'Excellent'))`. 
```{r}
brfss = 
  brfss_smart2010 |>
  as_tibble()

brfss = 
  brfss |>
  janitor :: clean_names() |>
  rename(state = locationabbr, county = locationdesc) |>
  filter(topic %in% 'Overall Health') |>
  mutate(response = fct_relevel(response, 
                                'Poor', 'Fair', 'Good', 'Very good', 'Excellent'))

brfss
```


In 2002, which states were observed at 7 or more locations?
In 2002, CT, FL, MA, NC, NJ, PA these 6 states were observed with 7 or more locations.

To know which states had observed 7 or more locations in 2002, we firstly need to use filter to filter our the 2002 year rows with `filter(year == '2002')`, then, we use `group_by(state)` function to get each state grouped together. Then we need to count the numbers of individual distinct county in each states and put the number into a new columns of `loc_num` by `summarise()`. Finally, we `filter()` the states with `loc_number >=7` to get the results we want, and `knitr::` the table. 
```{r}
brfss_2002 =
  brfss |>
  filter(year == '2002') |>
  group_by(state) |>
  summarise(loc_num = n_distinct(county)) |>
  filter( loc_num >= 7) |>
  knitr::kable()

brfss_2002
```

What about in 2010?
In 2010, CA, CO, FL, MA, MD, NC, NE, NJ, NY, OH, PA, SC, TX, and WA these 14 states were observed with 7 or more locations.

To know which states had observed 7 or more locations in 2010, we firstly need to use filter to filter our the 2010 year rows with `filter(year == '2010')`, then, we use `group_by(state)` function to get each state grouped together. Then we need to count the numbers of individual distinct county in each states and put the number into a new columns of `loc_num` by `summarise()`. Finally, we `filter()` the states with `loc_number >=7` to get the results we want, and `knitr::` the table. 
```{r}
brfss_2010 =
  brfss |>
  filter(year == '2010') |>
  group_by(state) |>
  summarise( loc_num = n_distinct(county))|>
  filter( loc_num >= 7) |>
  knitr::kable()

brfss_2010
```

Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. 

To construct the dataset that is limited to excellent responses, we use `filter()` function to focus on `response == 'Excellent'`. Then we use `select()` function to select the columns only with year, state, and data_value variables. To make the dataset more clean and easy to observed, we will use `group_by(year, state)` to help us to see the values grouped in same year and individual states. Finally, we use `summarize()` function to get the construct a new column of `mean_dv` to get the mean of `data_value`. Then we are ready to make the graph.
```{r}
brfss_dataset =
  brfss |>
  filter(response == 'Excellent') |>
  select(year, state, data_value) |>
  group_by(year, state) |>
  summarize(mean_dv = mean(data_value))

brfss_dataset
```

Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

It's time to make a graph. To make a spaghetti plot, we need to use `ggplot(aes(x = year, y = mean_dv, color = state, group = state))` to take in the value of `x = year`, `y = mean_dv`, and specify that we need to have different states in different colors with `color = state` and `group = state`. Since it is a spaghetti graph, we need to use `geom_line()` to plot and add title names and x, y variable names using `labs(title = 'Mean Data Value over Time', x = 'Year', y = 'Average Data Value')`. Our spaghetti plot is created and is save to pdf using `ggsave()`.
```{r}
brfss_spaghetti =
  brfss_dataset |>
  ggplot(aes(x = year, y = mean_dv, color = state, group = state)) + 
  geom_line() +
  labs(
    title = 'Mean Data Value over Time',
    x = 'Year',
    y = 'Average Data Value')

brfss_spaghetti

ggsave('brfss_spaghetti.pdf', height = 10, width = 12)
```
Comments:
From the graph, we can see that most of the states have mean data_value range from 16.5 to 27.5, and most of them have up and down values between 2002 to 2010. Some states like CA and CO had relatively high mean dv all the time around 28, and some states like WV have an extreme low mean dv at 2005 and 2007 (below 15). However, though with variation between the 8 years, the overall change of mean data_value is relatively small from 2002 to 2010.


Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

To make a two-panel plot among locations in NY state, we use `filter(state %in% 'NY')` to get NY rows selected, then we `filter()` for the years 2006 and 2010, we need to use `filter(year %in% c('2006','2010'))`. 
Now we are ready to plot the two-panel plot with `ggplot(aes(x = response, y = data_value))` to take in the value of `x = response`, `y = data_value`. Then we need to use `geom_boxplot()` to plot the boxplot which is clear to comparison and use `facet_grid(. ~ year)` to wrap the 1d to 2d. In this case, we are comparing 2006 and 2010 years so we wrap the variable year to make a side by box-plot. Then we add title names and x, y variable names using `labs(title = 'Data Value distribution over Response in NY for 2006 & 2010', x = 'Response', y = 'Data Value')`. Our two-panel box-plot is created and is save to pdf using `ggsave()`.
```{r}

brfss_box = 
  brfss |>
  filter(state %in% 'NY') |>
  filter(year %in% c('2006','2010')) |>
  ggplot(aes(x = response, y = data_value)) +
  geom_boxplot() +
  facet_grid(. ~ year) +
  labs(
    title = 'Data Value distribution over Response in NY for 2006 & 2010',
    x = 'Response',
    y = 'Data Value'
  )

brfss_box

ggsave('brfss_box.pdf', height = 10, width = 12)
```
Comments:
From the graph, we can see that in the two years 2006 and 2010, the response patterns are similar that good and very good responses have relatively high data value around 30-35, and excellent response is in the middle around 20-25, then the fair around 10-15 and lastly the poor having lowest data value below 5. Difference is in 2010, the very good response obtaining higher data value of around 35 dv along with high variation (longer box) and the good response having lower data value of around 32 dv with low variation (shorter box). Excellent is relatively the same with longer low data value as a tail in 2010, and fair response having lower variation (shorter box + shorter tail). 


Problem 3

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

First, we need to read and load the data by using `read_csv()` and we need to `skip = 4` to skip the first 4 rows, since these rows only contain sentences not data. Then we use `janitor::clean_names()` to clean the variable names. After that, we need to filter the rows to leave only `age>21` using `filter()`. Then we need to encode the data with reasonable variable class, we use `mutate()` to recode 1 to male and 2 to female. We need to do the same to education by recode 1 to `Less than high school`, 2 to `High school equivalent`, and 3 to `More than high school`. After recoding, we need to make them as factors by using `mutate( sex = as.factors(sex), education = as.factor(education))`. Finally, we omit the na data by using `drop_na()`.
```{r}
nhanes_covar =
  read_csv('nhanes_covar.csv', skip = 4) |>
  janitor::clean_names() |>
  filter(age > 21) |>
  mutate(
    sex = recode(sex, 
                 '1' = 'male',
                 '2' = 'female'),
    education = recode(education,
                       '1' = 'Less than high school',
                       '2' = 'High school equivalent',
                       '3' = 'More than high school'),
    sex = as.factor(sex),
    education = as.factor(education)) |>
  drop_na()

nhanes_covar
```

First, we need to read and load the `nhanes_accel.csv` data by using `read_csv()`. Then we use `janitor::clean_names()` to clean the variable names. After that, we need to use `pivot_longer()` to make the data clean and tidy with variable `Number` contain the values of `MIMS` in a column, so that it can shrink the data into fewer columns. It looks more clean and tidy.(minized the min1, min2 to min1440 with `min1:min1440`, and use `values_to = 'MIMS`, Finally, we omit the na data by using `drop_na()`.
```{r}
nhanes_accel = 
  read_csv('nhanes_accel.csv') |>
  janitor::clean_names() |>
  pivot_longer(
    min1:min1440,
    values_to = 'MIMS',
    names_to = 'Number',
    names_prefix = 'min') |>
  drop_na()

nhanes_accel
  
```


Joined the two table together. 

We joined the two table together using `left_join()` and the join key is `by = 'seqn'`.
```{r}
nhanes_joint = 
  left_join(nhanes_covar, nhanes_accel, by = 'seqn')

nhanes_joint
```


Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

In order to make a reader-friendly table, we use `group_by(sex, education)` to group the joined table in each category. Then, we are ready to make the plots. We use `ggplot(aes(x = age, y = education, fill = sex))` to load in the data in the plot. In this case, we need to do the comparison between different sex in different education level, so we are using `geom_boxplot()`, and we use `facet_grid()` to wrap the same sex into each three paralled boxes to make the graph to be compared. Finally, we use `labs()` to add title, and variable names.
```{r}
nhanes_table =
  nhanes_joint |>
  group_by(sex, education) 


age_plot = 
  nhanes_table |>
  ggplot(aes(x = age, y = education, fill = sex)) +
  geom_boxplot(alpha = 0.5) +
  facet_grid(sex~.) +
  labs(
    title = 'Age Distribution on different Sex across Education Level',
    x = 'Age',
    y = 'Education'
  )

age_plot
```
Comments: The overall pattern for borth male and female is that the education level lower, the mean age is higher for both sex. It makes sense, since people with older age when they are young back in the history, they are less likely to receive high education. While, younger people nowadays are in a modern society, so they are more likely to get higher education. This follows the trend of how society develops. On this boxplot, we can see both female and male groups have very similar distribution when the education level is less than high school. The female group age is slightly higher than the male group age. (mean around 61 in female and 60 in male) For the education level equivalent to high school, female group has mean age larger than the mean age of the male group age. The whole distribution is shifting to the higher age for female than male. (mean around 62 for female and around 54 for male) For the education level higher than high school, both female and male groups have similar age distribution, with female age mean lower than male mean age, and the male age has a wider distribution. (mean around 41 for female and around 45 for male) 


Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

In order to make the analysis, we need to make a clean table first by using `group_by()` to group the same sequence, education, sex, age together. Then we need to use the `summarise()` to get the sum of the same grouped sum of MIMS and make it into a new column of `total_activity`.
Then we are ready to make the plot by using the `nhanes_clean` table with `ggplot(aes(x = age, y = total_activity, color = sex))` to load in the data to the plot, and use `geom_point()` to deal with the discrete data distribution, and then use the `geom_smooth()` to better see the trend with the smooth line added. Since this case, we need to compare the different sex total activity levels in each education group, we are wraping the education into each group with `facet_grid(.~education)`. Finally, we are adding the title and x,y labels with `labs(title = 'Total activities over Age',x = 'Age', y = 'Total Activities')`.
```{r}
nhanes_clean =
  nhanes_table |>
  group_by(seqn, education, sex, age) |>
  summarise(
    total_activity = sum(MIMS)
  )

nhanes_clean

activity_plot =
  nhanes_clean |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(.~education) +
  labs(
    title = 'Total activities over Age',
    x = 'Age',
    y = 'Total Activities'
  )

activity_plot
```
Comments: The overall total activity pattern shows a downward correlation for both sex in each education levels. The decreasing trend is obvious that when age increased in both male and female groups, the total activities dropped. For the lower than high school education level, both male and female have a faster decreasing trend with male a little higher than female after age 40, and the total activity has a small peak around 60 for both sex. (Both dropped from 17500 to 8750) For the education level of high school, both male and female having a low point around 60 instead and male total activity level is almost always lower than female and stably decreasing from 15000 to 10000. While female total activities at first rise surprisingly to around 17500 and then drops sharply to 100000 and go to next small peak to 12500 and drops back to 100000. For the education level higher than high school, both female and male activities are much stable than the other groups and it drop from 13750 to around 100000, and male has a small peak around age 50 and female has a small peak later around 60. Overall, female has higher total activity than male. Overall, the education level may play a factor to slow down the total activity dropping rate. 


Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

Finally, we make the last plot using `nhanes_table`. We load data inside the plot with `ggplot(aes(x = Number, y = MIMS, group = seqn, color = sex))`. In this case, we are adding `seqn` as individual samples into the plot, so we need to have `group = seqn`, and then, we are adding points to the graph with `geom_point(alpha = 0.3)`. Then we need to see the trend more clearly by adding `geom_smooth(aes(group = sex), se = FALSE)`. Then we are grouping each education level together as the above plots with `facet_grid(.~education)`. Finally, we add the title and x,y labels with `labs()`.
```{r}
threepanel_plot =
    nhanes_table |>
  ggplot(aes(x = Number, y = MIMS, group = seqn, color = sex)) +
  geom_point(alpha = 0.3) +
  geom_smooth(aes(group = sex), se = FALSE) +
  facet_grid(.~education) +
  labs(
    title = '24-Hour Activities over Age for Education Level',
    x = 'Time',
    y = 'Total Activities'
  )

threepanel_plot

  
```
Comments: Overall pattern shows that both female and male have similar pattern for the education level below high school, and they both have a total activities around 25 to 35, and they both get to a lower point around the 11am at the 24 hours. While the more than high school education group has a different pattern that male have a high variation before the small lower point around 8-9am and another high point after the down time around 2-3pm. The second peak also involve some female individuals. Moreover, the lower than and equivalent to high school education groups have male activities dominated while the female activities dominated for the education level higher than high school. Therefore, we may get a clue that higher education may play some role with increasing male total activity level. 
